Finch: A Datastructure-Driven Array Programming Language

### 2024-03-19 main.tex
fix
Update on Overleaf.

\newcommand\teo[1]{\textcolor{red}{teo:#1}}
\newcommand\help[1]{\textcolor{green}{Help Needed:#1}}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
%
These frameworks have been the
subject of extensive industry investment to enable performant implementations,
and often operate at the peak capacity of the hardware they run on
\cite{lo_roofline_2015}.

The success and ubiquity of arrays is likely due to their simplicity. 
%
Since
their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that every entry the array
represents is stored faithfully at a location determined by a simple row or
column-major mapping from multidimensional space to linear memory.
%
This
simplicity enables extensive interoperability, convenience layers, and
optimizations by breaking the abstraction barrier between array representation
and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
loop-carry-dependencies, and more. Several optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.  However, while dense arrays are the easiest way to program, the
world is not all dense.

%Bring in structure
% - old arrays are Dense, rectilinear, integer grids
% - The world is not like that
% - Sparsity, runs of repeated values, symmetry
% - Lots of citations, perhaps a few figures from my powerpoint
%Sparse arrays
% - Networks (graphblas, graph frameworks)
% - Simulations (BLAS, symblas, banded blas, etc)
% - Databases (database engines)
%Run-length encoding
%Symmetry
%Bands
%Padding
%Blocks
\help{This section needs citations, please help expand it and make it more
convincing. A figure could be nice.} Our world is full of structured arrays.
Sparse arrays (which store only nonzero elements) describe networks and
simulations. Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)
Symmetry, bands, padding, and blocks are all natural matrix properties which
arise in the course of scientific computing or simulations of the natural world.
Operators, such as convolution, can be expressed as structured arrays. For
example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter.

\help{This section needs citations}
\textbf{Currently, support for structured data is fragmented and incomplete}.
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure and architecture to get performance.
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{datastructure
flexibility}. Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, or SciPy. However, libraries will only ever support a subset of
programs on a subset of datastructure combinations. Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semirings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values. While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays.  Special-purpose
compilers like TACO or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. This trade-off is
visualized in figure \ref{fig:datavsprogram}.

\begin{table}[h!]
  \help{This table needs citations, as well as a little bit of double checking}
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  High-Level API           &            & \checkmark &            &            & \checkmark \\
  Automatic API Fusion     &            &            &            &            & \checkmark \\
%  Parallelism              & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark & \checkmark \\
  Early Break              &            & \checkmark &            &   \checkmark         & \checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
  \end{table}

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
  Sparse                   &            & \checkmark &            &            & \checkmark \\
  Run-length               &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
  Ragged                   &            &            & \checkmark &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}

We see two main challenges to writing efficient code over structured data.

\textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory. For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value. These zero regions or repeated regions are optimization opportunities, and we must adapt the program to avoid repetitive work on these regions by referencing the stored structure.

\textbf{Performance is highly structure dependent}: Structure aware kernels are dramatically faster than unstructured ones, and the landscape of implementation decisions is dramatically unpredictable. For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.

In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code. In this model, performance engineers can more efficiently search the
complex landscape of programs and datastructures to find the best implementation. In this programming model, we can express certain concepts in programs, and others in data. For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.

We develop a new array programming language, Finch, which supports \textit{both}
flexible programming constructs and diverse data structures. Finch supports a
familiar programming language of loops, statements, if conditions, breaks, etc,
over a wide variety of array structures, such as sparsity, run-length-encoding,
symmetry, triangles, padding, or blocks. The Finch compiler uses the structure
of the data to generate efficient implementations of these programs.
%It does this using Looplets, a new language that decomposes datastructure iterators hierarchically. Looplets represent the control-flow structures needed to iterate over them simultaneously with other structures in the program. Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc. can be implemented using simple compiler passes like term rewriting and constant propagation.

\subsection{Contributions}

\begin{enumerate}
\item More complex array structures than ever before. A complete level-by-level
structure-description language for expressing the structure of data
hierarchically. The first such set of formats to efficiently capture banded,
triangular, run-length-encoded, or sparse datasets, and any combination thereof.
\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. To our knowledge, the Finch programming language is the first
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather.
\item A compiler that specializes programs to data structures 
predictably, facilitating an expressive language that makes it easier to search the complex
space of algorithms and datastructures. It is easy to implement new datastructures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
initialization or finalization.
\item We evaluate the productivity of our language in several case studies,
showing that Finch can be used to accelerate a wide range of applications, 
from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

\section{Background}
\subsection{Looplets}
\subsection{FiberTrees}

\subsection{Concordant Iteration}

\subsection{Protocols}

\section{Bridging Looplets and Finch: The Tensor Interface}
\teo{I am doing stuff in this section.}
\input{sections/tensors.tex}

\subsection{Wrapper Tensors}

\subsection{Scalars}

\subsubsection{Sparse Scalars}
\subsubsection{Early Break Scalars}

\section{The Finch Language}

\subsection{Syntax and Semantics}

\section{The Finch Compiler}

\subsection{Dimensionalization}

\subsection{Concordization}

\subsection{Bounds Analysis}

\subsection{Performance Warnings}

\subsection{Wrapperization}

\subsection{Simplification and Algebraic Transformations}

\section{Evaluation}

\subsection{Data-Driven Performance Engineering}
\subsubsection{Sparse-Sparse Matrix Multiply}

\begin{figure}
	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (4 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension. Note that inner products necessarily requires $O(n^2)$ work and taco's outer products format is dense. Finch can use a sparse outer products format and thus has an asymptotic advantage that becomes evident as the output dimensions grow.}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of gustavson's algorithm between Finch and Taco on some larger matrices}
\end{figure}

Examples that demonstrate performance engineering in a datastructure-driven model

\subsubsection{SpMV}
Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data. More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats. Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix. % I'm not sure if the right place for this is here - maybe should come earlier?

Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns. Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.

The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks. The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats. In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.

We compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface [cite], Ahrens et al. to test a variable block row format partitioning strategy [cite], and Kjolstad et al. to test the TACO library [cite]. In addition, we also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases. The dense vector is randomly generated. We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures 1 and 2.

% TODO: make this be 3 columns
\begin{figure}
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[i] += A[i, j] * x[j]
                end
                return y
            end
        \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[j] += A[i, j] * x[i]
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \begin{minipage}[t]{0.36\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _
                    let x_j = x[j]
                        y_j .= 0
                        for i = _
                            let A_ij = A[i, j]
                                y[i] += x_j * A_ij
                                y_j[] += A_ij * x[i]
                            end
                        end
                        y[j] += y_j[] + diag[j] * x_j
                    end
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \caption{Performance of SpMV across various tools.}
\end{figure}

\subsubsection{Tensor Formats}
We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix. To represent scalar values, we benchmarked with both the Element and Pattern levels. The SparseList, SparseVBL, and SparseBand formats were functional for all matrices, but the SparsePoint format was applicable only for the synthetic permutation matrices. The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.

We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor. Namely, matrices with a clear blocked structure like $exdata\_1$, $TSOPF\_RS\_b678\_c1$, and $heart3$ performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation. Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.

There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean. The matrices in the SNAP collection which represent graph datasets are boolean and had an average speedup of X over TACO and X over the same Finch program, but with an Element format.

\subsubsection{Symmetric SpMV}
Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix. We can apply this strategy with any level format. Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry. However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage. Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO . Notably, there is a 1.9x speedup for the HB/saylr4 matrix. 
\subsubsection{4D Blocked SpMV}

%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png

\begin{figure}
    \begin{minipage}[t]{0.5\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \includegraphics[width=\linewidth]{spmv_performance_sorted_(faster_than_taco).png}
    \end{minipage}%
    \begin{minipage}[t]{0.5\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \includegraphics[width=\linewidth]{spmv_performance_sorted_(slower_than_taco).png}
    \end{minipage}
    \caption{Performance of SpMV across various tools.}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
\end{figure}

\subsection{Programming over flexible data}

\subsubsection{Image Morphology}

\begin{figure}
	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (2 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (4 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{hist_speedup_over_opencv.png}
    \caption{Performance of Finch on masked histogram task.}
\end{figure}

\subsubsection{Graph Analytics}
\begin{figure}
	\includegraphics[width=\linewidth]{bfs_speedup_over_graphs.jl.png}
	\includegraphics[width=\linewidth]{bellmanford_speedup_over_graphs.jl.png}
    \caption{Performance of graph apps across various tools.}
\end{figure}

Push-Pull BFS in Finch:
\begin{minted}{julia}
V = Tensor(Dense(Element(false)))
P = Tensor(Dense(Element(0)))
F = Tensor(SparseByteMap(Pattern()))
_F = Tensor(SparseByteMap(Pattern()))
A = Tensor(Dense(SparseList(Pattern())))
AT = Tensor(Dense(SparseList(Pattern())))

function finch_bfs_push_kernel(_F, F, A, V, P)
    @finch begin
        _F .= false
        for j=_, k=_
            if F[j] && A[k, j] && !(V[k])
                _F[k] |= true
                P[k] <<choose(0)>>= j #Only set the parent for this vertex
            end
        end
        return _F
    end
end

function finch_bfs_pull_kernel(_F, F, AT, V, P)
    p = ShortCircuitScalar{0}()
    @finch begin
        _F .= false
        for k=_
            if !V[k]
                p .= 0
                for j=_
                    if F[follow(j)] && AT[j, k]
                        p[] <<choose(0)>>= j #Only set the parent for this vertex
                    end
                end
                if p[] != 0
                    _F[k] |= true
                    P[k] = p[]
                end
            end
        end
        return _F
    end
end)
\end{minted}

\subsection{Implementing Numpy Array API in Finch}
\subsubsection{The Finch High-Level API (Needs a Name)}

\subsubsection{Finch Logic}

\subsubsection{Finch Interpreter}

\subsubsection{Lowering}
\subsubsection{Heuristic Optimization}

Find an example where fusing the python interface gives a big speedup over non-fused kernels.

%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
%in-place.
%dot((v^t .* u), w)) vs. 
%(v^t .* dot(u, w))

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
    To Mateusz, Hameer, and Jaeyeon for their excellent programming contributions to the Finch codebase.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{FinchOOPSLA.bib}

>>>>

\newcommand\teo[1]{\textcolor{red}{teo: #1}}
\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}

@@

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle
\input{sections/intro}
\input{sections/background}
\input{sections/lang}
\input{sections/tensors}
\input{sections/compiler}
\input{sections/eval}
%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
    To Mateusz, Hameer, and Jaeyeon for their excellent programming contributions to the Finch codebase.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{FinchOOPSLA.bib, FinchOOPSLAOverleaf.bib}
